Concurrent Programming in Harmony

### 2021-12-24 book/paper.tex
Pass over Ch 18

Tony Hoare, who came up with the concept of split binary semaphores, devised
an abstraction of the concept in a programming language paradigm called
\emph{monitors}~\cite{Hoare74}.
\index{monitor}%
(A similar construct was independently invented by Per Brinch Hansen~\cite{BH73}.)
A monitor is a special version of an object-oriented \emph{class}, comprising
a set of variables and methods that operate on those variables.
A monitor also has special variables called \emph{condition variables},
one per waiting condition.
There are two operations on condition variables: <{wait}>
\index{wait}%
and
\texttt{signal}.
\index{signal}%

Harmony does not have language support for Hoare monitors but it
has a module called \texttt{hoare}.
\autoref{fig:hoare} shows its implementation.
A monitor uses a hidden split binary semaphore.
The mutex semaphore is acquired when entering a monitor and released
upon exit.
Each condition variable maintains a binary semaphore and a counter for
the number of threads waiting on the condition.
<{wait}> increments the condition's counter, releases the monitor mutex,
blocks while trying to acquire the condition's semaphore, and upon resuming
decrements the counter---in much the same way as we have seen for split binary
semaphores (SBS).
\texttt{signal} checks to see if the condition's count is non-zero, if so
releases the condition's semaphore,
and then blocks by trying to acquire the mutex again.

\autoref{fig:hoaretest} presents a bounded buffer implemented using
Hoare monitors.
It is written in much the same way you would if using
the SBS technique (see \autoref{ex:bbsbs}).
However, there is no \texttt{release\_one} method.
Instead, one can conclude
that <{put}> guarantees that the queue will be non-empty, and
\texttt{signal} will check if there are any threads waiting for
this event.  If so, \texttt{signal} will pass control to one such thread
and, unlike \texttt{release\_one}, re-enter the critical
section afterwards by acquiring the <{mutex}>.

Implementing a reader/writer lock with Hoare monitors is not quite so
straightforward, unfortunately.  When a writer releases the lock, it
has to choose whether to signal a reader or another writer.  For that
it needs to know if there is a reader or writer waiting.  The simplest
solution would be to peek at the counters inside the respective condition
variables, but that breaks the abstraction.  The alternative is for the
reader/writer implementation to keep track of that state explicitly,
which complicates the code.  Also, it requires a deep understanding of
the SBS method to remember to place a call to \texttt{signal} in the
\texttt{read\_acquire} method that releases additional readers that
may be waiting to acquire the lock.

@@

As in Hoare monitors, there is a hidden mutex associated with each Mesa monitor,
and the mutex must be acquired upon entry to a method and released
upon exit.
Mesa monitors also have condition variables that a thread can wait on.
Like in Hoare monitors, the <{wait}> operation releases the mutex.
The most important difference is in what \texttt{signal} does.
To make the distinction more clear, we shall call the corresponding Mesa
operation <{notify}> rather than \texttt{signal}.
\index{notify}%
When a thread <{p}> invokes <{notify}>, it does not immediately pass 
control to a thread that is waiting on the corresponding condition (if there
is such a thread).  Instead, <{p}> remains in the critical section
until it leaves the monitor explicitly
(by calling either <{release}> or <{wait}>).
At that point, any thread that
was notified will have a chance to enter the critical section, but they compete
with other threads trying to enter the critical section.

@@

On the right is a Mesa monitor.  The only way into the bathroom is through
the hall.  When thread <{p}> invokes \textbf{notify} on a bedroom, it
selects a thread <{q}> in that bedroom, if any.
Thread <{q}> goes into the hall,
joining any other threads that are also waiting to
enter the bathroom.
If there are no threads in that bedroom, \textbf{notify} is a no-op.
Importantly, in either case <{p}> remains in the bathroom.
Finally, when <{p}> leaves the bathroom, one of the
threads in the hall can enter the bathroom.  But that may not be <{q}>.
Assuming every thread eventually leaves the bathroom and the system is
fair, eventually <{q}> will enter the bathroom.  However, it is no longer
guaranteed that the condition it was waiting for still holds.
Therefore, the first thing <{q}> must
do is check, again, to see if the condition holds.  If not, it should go back
into the bedroom corresponding to the condition.
\end{comment}

@@

In Harmony, a bag is usually represented by a dictionary that maps
the elements of the bag to their multiplicities.
For example, the value \texttt{\{ .a: 2, .b: 3 \}} represents a bag
with two copies of \texttt{.a} and three copies of \texttt{.b}.
The \texttt{bag} module (\autoref{ap:bag}) contains a variety of handy
functions on bags.

<{wait}> adds the context of the thread to the bag.
This increments the number of threads in the bag with the same context.
<{wait}> then loops until that count is restored to the
value that it had upon entry to <{wait}>.
<{notify}> removes an arbitrary context from
the bag, allowing one of the threads with that context to
resume and re-acquire the lock associated with the monitor.
<{notifyAll}> empties out the entire bag, allowing all threads
in the bag to resume.

@@

Note that <{wait}> is always invoked within a <{while}>
loop that checks for the condition that the thread is waiting for.
It is \emph{imperative} that there is always a <{while}> loop
around any invocation of <{wait}> containing the negation of
the condition that the thread is waiting for.  Many implementation
of Mesa condition variables depend on this, and optimized implementations
of condition variables often allow so-called ``spurious wakeups,''
where <{wait}> may sometimes return even if the conditon
variable has not been notified.
As a rule of thumb, one should always be able to replace <{wait}>
by <{unlock}> followed by <{lock}>.
This turns the solution into a busy-waiting one, inefficient but still
correct.

@@

% While Hoare moniters embody the split binary semaphore approach,
% Mesa monitors are much closer to the busy-waiting approach.
When using Mesa condition variable you have to be careful to invoke
\textbf{notify} or \textbf{notifyAll} in the right places.
Much of the complexity of programming with Mesa condition variables
is in figuring out when to invoke <{notify}> and when to invoke
<{notifyAll}>.  As a rule of thumb: be conservative---it is
better to wake up too many threads than too few.
In case of doubt, use <{notifyAll}>.
Waking up too
many threads may lead to some inefficiency, but waking up too few
may cause the application to get stuck.  Harmony can be particularly
helpful here, as it examines each and every corner case.
You can try to replace each <{notifyAll}> with <{notify}> and see
if every possible execution of the application still terminates.

>>>>

Tony Hoare, who came up with the concept of split binary semaphores (SBS),
devised an abstraction of the concept in a programming language paradigm
called \emph{monitors}~\cite{Hoare74}.
\index{monitor}%
(A similar construct was independently invented by Per Brinch Hansen~\cite{BH73}.)
A monitor is a special version of an object-oriented \emph{class}, comprising
a set of variables and methods that operate on those variables.
A monitor also has special variables called \emph{condition variables},
one per waiting condition.
There are two operations on condition variables: <{wait}>
\index{wait}%
and
<{signal}>.
\index{signal}%

Harmony does not have language support for monitors, but it
has a module called \texttt{hoare}.
\autoref{fig:hoare} shows its implementation.
A Hoare monitor uses a hidden split binary semaphore.
The mutex semaphore is acquired when entering a monitor and released
upon exit.
Each condition variable maintains a binary semaphore and a counter for
the number of threads waiting on the condition.
Method <{wait}> increments the condition's counter, releases the monitor mutex,
blocks while trying to acquire the condition's semaphore, and upon resuming
decrements the counter---in much the same way as we have seen for SBS.
Method <{signal}> checks to see if the condition's count is non-zero, if so
releases the condition's semaphore,
and then blocks by trying to acquire the mutex again.

\autoref{fig:hoaretest} presents a bounded buffer implemented using
Hoare monitors.
It is written in much the same way you would if using
the SBS technique (see \autoref{ex:bbsbs}).
However, there is no \texttt{release\_one} method.
Instead, one can conclude
that <{put}> guarantees that the queue will be non-empty, and
<{signal}> will check if there are any threads waiting for
this event.  If so, <{signal}> will pass control to one such thread
and, unlike \texttt{release\_one}, re-enter the critical
section afterwards by acquiring the <{mutex}>.

Implementing a reader/writer lock with Hoare monitors is not quite so
straightforward, unfortunately.  When a writer releases the lock, it
has to choose whether to signal a reader or another writer.  For that
it needs to know if there is a reader or writer waiting.  The simplest
solution would be to peek at the counters inside the respective condition
variables, but that breaks the abstraction.  The alternative is for the
reader/writer implementation to keep track of that state explicitly,
which complicates the code.  Also, it requires a deep understanding of
the SBS method to remember to place a call to <{signal}> in the
\texttt{read\_acquire} method that releases additional readers that
may be waiting to acquire the lock.

@@

As in Hoare monitors, there is a hidden mutex associated with each Mesa monitor,
and the mutex must be acquired upon entry to a method and released
upon exit.
Mesa monitors also have condition variables that a thread can wait on.
Like in Hoare monitors, the <{wait}> operation releases the mutex.
The most important difference is in what <{signal}> does.
To make the distinction more clear, we shall call the corresponding Mesa
operation <{notify}> rather than <{signal}>.
\index{notify}%
Unlike <{signal}>, when a thread <{p}> invokes <{notify}>
it does not immediately pass 
control to a thread that is waiting on the corresponding condition (if there
is such a thread).  Instead, <{p}> continues executing in the critical section
until it leaves the monitor (by calling <{release}>) or
releases the monitor (by calling <{wait}>).
Either way, any thread that was notified will now have a chance to enter
the critical section, but they compete
with other threads trying to enter the critical section.

@@

On the right is a Mesa monitor.  The only way into the bathroom is through
the hall.  When thread <{p}> invokes <{notify}> on a bedroom, it
selects a thread <{q}> in that bedroom, if any.
Thread <{q}> goes into the hall,
joining any other threads that are also waiting to
enter the bathroom.
If there are no threads in that bedroom, <{notify}> is a no-op.
Importantly, in either case <{p}> remains in the bathroom.
Finally, when <{p}> leaves the bathroom, one of the
threads in the hall can enter the bathroom.  But that may not be <{q}>.
Assuming every thread eventually leaves the bathroom and the system is
fair, eventually <{q}> will enter the bathroom.  However, it is no longer
guaranteed that the condition it was waiting for still holds.
Therefore, the first thing <{q}> must
do is check, again, to see if the condition holds.  If not, it should go back
into the bedroom corresponding to the condition.
\end{comment}

@@

In Harmony, a bag is usually represented by a dictionary that maps
the elements of the bag to their multiplicities.
For example, the value <{{ .a: 2, .b: 3 }}> represents a bag
with two copies of <{.a}> and three copies of <{.b}>.
The <{bag}> module (\autoref{ap:bag}) contains a variety of handy
functions on bags.

Method <{wait}> adds the context of the thread to the bag.
This increments the number of threads in the bag with the same context.
<{wait}> then loops until that count is restored to the
value that it had upon entry to <{wait}>.
Method <{notify}> removes an arbitrary context from
the bag, allowing one of the threads with that context to
resume and re-acquire the lock associated with the monitor.
<{notifyAll}> empties out the entire bag, allowing all threads
in the bag to resume.

@@

Note that <{wait}> is always invoked within a <{while}>
loop that checks for the condition that the thread is waiting for.
It is \emph{imperative} that there is always a <{while}> loop
around any invocation of <{wait}> containing the negation of
the condition that the thread is waiting for.  Many implementation
of Mesa condition variables depend on this, and optimized implementations
of condition variables often allow so-called ``spurious wakeups,''
where <{wait}> may sometimes return even if the conditon
variable has not been notified.
As a rule of thumb, one should always be able to replace <{wait}>
by <{release}> followed by <{acquirire}>.
This turns the solution into a busy-waiting one, inefficient but still
correct.

@@

% While Hoare moniters embody the split binary semaphore approach,
% Mesa monitors are much closer to the busy-waiting approach.
When using Mesa condition variables, you have to be careful to invoke
<{notify}> or <{notifyAll}> in the right places.
Much of the complexity of programming with Mesa condition variables
is in figuring out when to invoke <{notify}> and when to invoke
<{notifyAll}>.  As a rule of thumb: be conservative---it is
better to wake up too many threads than too few.
In case of doubt, use <{notifyAll}>.
Waking up too
many threads may lead to some inefficiency, but waking up too few
may cause the application to get stuck.  Harmony can be particularly
helpful here, as it examines each and every corner case.
You can try to replace each <{notifyAll}> with <{notify}> and see
if every possible execution of the application still terminates.
